# Generalizability Theory

```{r}
# Packages used:
library(psych)
library(MPsychoR)
data("Rmotivation")

# Create data frame with only Hybrid Motivation items.
HybMot <- subset(Rmotivation, 
                        select = grep("hyb", names(Rmotivation)))
# Remove rows  with any missing data.
HybMot <- na.omit(HybMot)
library("reshape2")
# Add person variable
Hyb1 <- data.frame(HybMot, person = 1:nrow(HybMot))
Hyblong <- melt(Hyb1, id.vars = c("person"), 
                variable.name = "item")
Hyblong$person <- as.factor(Hyblong$person)
summary(aov(value ~ person + item, data = Hyblong))
```

From this output we can calculate Cronbach's $\alpha$ with the following values:

```{r}
MSp <- 0.85
MSe <- 0.15

alpha <- (MSp - MSe)/MSp
print(alpha, digits = 2)
```

```{r}
ICC(HybMot)
```


```{r}
icchyb <- ICC(HybMot)
sqrt((0.85-0.15)/19)
sqrt((31.88-0.15)/777)

library("lme4")
VarCorr(lmer(value ~ (1|person) + (1|item), data = Hyblong))

library("gtheory")
gfit <- gstudy(data = Hyblong, formula = value ~ (1|person) + (1|item))
dfit <- dstudy(gfit, colname.objects = "person", colname.scores = "value", 
               data = Hyblong)
round(dfit$generalizability, 3)
```

### Multiple Sources of Error

Generalizability theory acknowledges that multiple sources of error can impact scores simultaneously, and allow for estimating the effects of each [@raykov2011introduction].
These various sources of error, or facets (e.g., items, raters, measurement occasions).
All measurements of behavior are conceptualized as being sampled from a *universe* of admissible observations [@raykov2011introduction].
If the observed score is expected to vary across a facet (e.g. vary across occasions, or vary depending on the items included, or the rater scoring), then that facet is a defining characteristic of the universe.
The idea of reliability is replace with the idea of generalizability, which ,instead of asking how accurately observed scores can reflect the true score (CTT), generalizability theory asks how accurately observed scores allow us to generalize about behavior of an individual in a particular universe.

Below is the code from the @mair2018modern text.
```{r}
library(gtheory)

data("Lakes")
phydat <- subset(Lakes, subtest == "physical")
phydat$item <- droplevels(phydat$item)
head(phydat)

formula <- score ~ (1|personID) + (1|raterID) + (1|item) + 
  (1|personID:raterID) + (1|personID:item) + (1|raterID:item)
gfit <- gstudy(formula = formula, data = phydat)
gfit
```

```{r}
dfit <- dstudy(gfit, colname.objects = "personID", colname.scores = "score", 
               data = phydat)
dfit$components
```

```{r}
dfit$var.error.abs
dfit$sem.abs
dfit$var.error.rel
dfit$sem.rel
dfit$dependability
dfit$generalizability
```

## Additional Readings

For more information of G theory, see @raykov2011introduction. 
For an example using the R package `lavaan` with G theory, see @Jorgensen2021Howestimateabsolute.